---
title: "Introduction to the `NHSRplotthedots` package"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Introduction to the `NHSRplotthedots` package}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the NHS-R community's collaborative package for building a specific type of statistical process control (SPC) chart, the XmR chart.
We are aiming to support the NHS England's
    ['Making Data Count' programme][mdc],
    a programme which encourages boards, managers, and analyst teams to present
    data in ways that show change over time, and drive better understanding of
    indicators than 'RAG' (red, amber, green) rated board reports often support.

This tutorial is a concise take on applying the function to some Accident and Emergency breach data for some NHS hospitals.
We'll take this from the `NHSRdatasets` package, another of our packages providing example datasets for training in an NHS/healthcare context.

The analyses below use the `tidyverse` packages: `dplyr` to manipulate the data, and `ggplot2` for plotting (when not using `NHSRplotthedots`).

Firstly, let's load the data, filter for University Hospital Birmingham NHS Foundation Trust, because they are a good example for the type of charts below.
Let's also do a simple timeseries plot to see why this is the case.

```{r dataset, message=FALSE, warning=FALSE}
library(NHSRplotthedots)
library(NHSRdatasets)
library(dplyr)
library(ggplot2)
library(scales)

data("ae_attendances")

ae_attendances %>%
  filter(org_code == "RRK", type == 1) %>%
  ggplot(aes(x = period, y = breaches)) +
  geom_point() +
  geom_line() +
  scale_y_continuous("4-hour target breaches", labels = comma) +
  scale_x_date("Date") +
  labs(title = "Example plot of A&E breaches for organsiation: 'RRK'") +
  theme_minimal()
```

This minimal plot shows the changes over time, and we will look at it in two ways.
The first, we'll look at 2016/17 & 2017/18, which is apparently stable and, during 2018, the trust merged with another large 3-hospital provider trust, so the numbers through A&E shot up dramatically when combined under one trust code.
We can use a change point in our plots to consider this.

Let's now use the `ptd_spc` function to draw our plot.

We need to provide it with a `data.frame` that contains our data, and provide a `value_field` for the y-axis, and a `date_field` for the x-axis.
In addition, to ensure the points are coloured correctly, we need to set the `improvement_direction` to "decrease", here (as fewer A&E breaches is better).

## Stable period

```{r stableperiod}
stable_set <- ae_attendances %>%
  filter(
    org_code == "RRK",
    type == 1,
    period < as.Date("2018-04-01")
  )

ptd_spc(
  stable_set,
  value_field = breaches,
  date_field = period,
  improvement_direction = "decrease"
)
```

From the chart above, we can see the centre line (mean) and the control limits calculated according the the XmR chart rules.
Our points that are between the control limits are within control and showing 'common-cause' or 'natural' variation.
We can see 8 sequential points coloured in blue. This is triggered by a rule that looks for >=7 points on one side of the mean.
This could be viewed as a period where fewer breaches than average were detected, which may hold some learning value for the organisation.

## Change point

From the first plot above, we can see that the change is noticed at 01/07/2019.
We understand the reason for the change (the merging of two trusts), and believe this will be the new normal range for the process, so it would be valid to rebase at this point.
To rebase (change the modelling period for control limits), we can pass a vector of dates at the point we wish to rebase using the `ptd_rebase()` function.

```{r changepoint}
change_set <- ae_attendances %>%
  filter(org_code == "RRK", type == 1)

ptd_spc(change_set,
  value_field = breaches,
  date_field = period,
  improvement_direction = "decrease",
  rebase = ptd_rebase(as.Date("2018-07-01"))
)
```

You can see that our limit calculation has now shifted at the rebase point.

## Faceting

In `{ggplot2}`, a `facet` refers to splitting or plotting by variables, into separate plots or regions.
Imagine you want to do an SPC for all trusts in a dataset, or all specialties at a trust.
Facet will help you do this, and you can pass the field that controls this to the `ptd_spc` argument `facet_field`.

Here we will pick 6 trusts at random and plot their breaches, faceting across them.

We'll also combine this with a couple more options that are specific to the plot.
This is done with the `ptd_create_ggplot` function in the backend, but also works with the generic `plot`.
To find out the available options, see the help-file for: `?ptd_create_ggplot`.

We will let the x-axis scale change for each plot for each plot (otherwise some would look crushed against a larger trusts) and we will also change the x-axis breaks to 3-months, as 1 month looked too busy.)

```{r facetvignette}
facet_set <- ae_attendances %>%
  filter(
    org_code %in% c("RRK", "RJC", "RJ7", "R1K", "R1H", "RQM"),
    type == 1,
    period < as.Date("2018-04-01")
  )

ptd_spc(
  facet_set,
  value_field = breaches,
  date_field = period,
  facet_field = org_code,
  improvement_direction = "decrease"
) %>%
  plot(fixed_y_axis_multiple = FALSE, x_axis_breaks = "3 months")
```

From this, the point-size seems too big, and it would be worth replacing the y-axis title with something better.
Again, this is done with the `plot` function arguments:

```{r facetvignette2}
facet_set <- ae_attendances %>%
  filter(
    org_code %in% c("RRK", "RJC", "RJ7", "R1K", "R1H", "RQM"),
    type == 1,
    period < as.Date("2018-04-01")
  )

ptd_spc(
  facet_set,
  value_field = breaches,
  date_field = period,
  facet_field = org_code,
  improvement_direction = "decrease"
) %>%
  plot(
    fixed_y_axis_multiple = FALSE,
    x_axis_breaks = "3 months",
    point_size = 2,
    y_axis_label = "Number of 4-hour A&E target breaches"
  )
```

As the plots are based on `{ggplot2}` you can override elements of the styling and theme as additional arguments using `theme`, or by using `theme_override` with your new theme elements in a `list`.

```{r facetvignette3}
facet_set <- ae_attendances %>%
  filter(
    org_code %in% c("RRK", "RJC", "RJ7", "R1K", "R1H", "RQM"),
    type == 1,
    period < as.Date("2018-04-01")
  )

a <- ptd_spc(
  facet_set,
  value_field = breaches,
  date_field = period,
  facet_field = org_code,
  improvement_direction = "decrease"
) %>%
  plot(
    fixed_y_axis_multiple = FALSE,
    x_axis_breaks = "3 months",
    point_size = 2,
    y_axis_label = "Number of 4-hour A&E target breaches"
  )

a + theme(axis.text.x = element_text(size = 6, angle = 45))
```

### Interactive plots

It's also possible to create interactive plots using the `{plotly}` package by replacing the call to `plot` with `ptd_create_plotly`.
This takes the same arguments as `plot` (and `ptd_create_ggplot`).

For instance, we can use the code from the last example to make an interactive version:

```{r plotly}
ptd_spc(
  facet_set,
  value_field = breaches,
  date_field = period,
  facet_field = org_code,
  improvement_direction = "decrease"
) %>%
  ptd_create_plotly(
    fixed_y_axis_multiple = FALSE,
    x_axis_breaks = "3 months",
    point_size = 2,
    y_axis_label = "Number of 4-hour A&E target breaches",
    icons_position = "none"
  )
```

## Come and join us!

This is a collaborative project that is still early in its life.
All contributors are volunteers, and more volunteers would be very welcome (there's even stuff that doesn't require `R` coding to help with!).

Find out more at:
[github.com/nhs-r-community/NHSRplotthedots][ptd_gh]

[mdc]: https://www.england.nhs.uk/publication/making-data-count/
[ptd_gh]: https://github.com/nhs-r-community/NHSRplotthedots
