---
title: "Introduction to the `NHSRplotthedots` package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the `NHSRplotthedots` package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(NHSRplotthedots)
```

___This vignette is a work in progress, feel free to contribute on NHS-R GitHub___

Welcome to the NHS-R community's collaborative package for building specific types of statistical process control charts.  We are aiming to support the NHS Improvement 'Making Data Count' programme, https://www.england.nhs.uk/a-focus-on-staff-health-and-wellbeing/publications-and-resources/making-data-count/, which aims to encourage boards, managers and analyst teams to present data in ways that present changes over time.

This tutorial is a concise take on applying the function to some Accident and Emergency breach data for a hospital.


```{r dataset}
library(NHSRdatasets)

data("ae_attendances")


ae_attedances %>% 
  filter(trust_code)

```
